Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-04 Manuscript/manuscript.Rmd
wip throuth figure 2

We created a resolution-specific response variable that indicated whether a genomic bin of unit width overlapped with a flanked TAD boundary ($Y=1$), or not ($Y=0$). The unit width of each bin on the linear genome corresponded to the resolution of the contact matrix on which domains were called. Moreover, TAD boundaries were flanked on either side by the same unit length as the genomic bins to account for uncertainty in boundary location. Expectedly, the median number of genomic bins, TADs called, and TAD boundaries increased with the increased resolution of cell-line specific Hi-C data (Table 1; Supplementary Tables 3). This correlation [?? What correlation ??] with resolution was found to be significant, as evidenced by large intraclass correlation coefficients (ICC) (Table 2; Supplementary Tables 4). However, the class imbalance problem, caused by the sparsity of TAD boundaries throughout the genome, was found to be much more consistant across resolutions in both cell lines. This imbalance was evidenced by low imbalance ratios, defined as the ratio of the number of genomic bins that overlapped with flanked TAD boundaries, to the much larger number of genomic bins that did not. This relationship was not significantly correlated with resoltuion for either cell line, indicating that there existed a severe disparity between classes at all resolutions of Hi-C data.

**Table 1. Domain Data Summaries and Imbalance Ratios.** For GM12878, average number of genomic bins, TADs called, and the corresponding boundaries and imbalance ratios, across each of the autosomal chromosomes (omitting chromosome 9), stratified by resolution. The resolution of the Hi-C contact matrix dictated the width of each genomic bin, with lower resolutions corresponding to wider genomic bins.

@@

**Table 2. Inter-Resolution ICCs.** For GM12878, two-way intraclass correlation coefficients between resolutions for the following quantities: Number of genomic bins, number of TADs called, number of unique TAD boundaries, and imbalance ratio. The ICC was computed across 21 of the autosomal chromosomes. P-values indicate the level of significance of the correlation between data resolution and each quantity.[??? What is ICC? !!!editted]

| Quantity                 | Inter-Resolution | 95% CI          | P-Value  |
|--------------------------|------------------|-----------------|----------|
| Number of Genomic Bins   | 0.464            | [0.244, 0.690]  | < 0.0001 |
| Number of TADs Called    | 0.627            | [0.426, 0.802]  | < 0.0001 |
| Number of TAD Boundaries | 0.611            | [0.407, 0.791]  | < 0.0001 |
| Imbalanced Ratio         | 0.114            | [-0.058, 0.372] | 0.109    |

<!--
## HIGH JACCARD INDICES IMPLY COLOCALIZATION OF MANY GROUP-SPECIFC ANNOTATIONS 
-->

## ESTABLISHING THE FEATURE SPACE USING INCREASINGLY GRANULAR RELATIONSHIPS BETWEEN FUNCTIONAL ANNOTATIONS AND GENOMIC BINS

The influences of TAD boundaries were assessed using functional genomic annotations, derived by ChIP-seq peak algorithms. Specific groups of annotations were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] and included chromatin state segementations, histone modifications, and transcription factor binding sites (TFBS).

@@

To establish the feature space for all downstream models, we constructed 3 different predictor-types. These predictor-type specific annotations represented the data matrix, $\textbf{X} = \{X_1, X_2, ..., X_p\}$, where $p$ was the complete set of functional annotations [?? $p$ is actually the total number of functional annotations, $X_i$ is the functional annotation itself ??]. Overlap counts was the least granular [?? Maybe define granular ??] predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined [?? what is refined in this context ??] predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C). A $log_{2}$ transformation was used as a normalization technique in order to account for the skewness present in the distance densities (Supplementary Figures 1-2). [?? Why would skewness matter, why not use raw data ??]

![**Predictor types.** Three separate types of predictors were considered when quantifying the association between TAD boundaries and genomic annotations. Each predictor type was used to build the feature space of specific models in the ensemble framework. (A) _Overlap Counts_ (OC) consider the total number of genomic features that overlapped with each genomic bin. (B) _Overlap Percent_ (OP) was calculated by dividing the sum of all feature widths within each specific bin by the total bin width. (C) _Distance_ ($log_{2}$ transformed) was calculated by measuring the distance (in bases) from the center of each genomic bin to the center of the nearest region defining a genomic annotation. ](C:/Users/stili/Documents/TAD_predictive_modeling/Manuscript/Figures/figure2_predictor_types.png)

@@

![**Figure 3. Ensemble Framework.** A diagram of the cell line specific ensemble framework detailing the combination of inputs used in the model building pipeline. The minoirty and majority classes were represented by the genomic bins that overlapped with flanked TAD boundaries or did not, respectively. Functional genomic annotations obtained from the UCSC Genome Browser Database composed the feature space ($\textbf{X} = \{X_1, X_2, ..., X_p\}$). A feature engineering step was included to isolate the predictor-type specific spatial relationship between annotations and genomic bins in increasing granularity, denoted by the red gradient. The data was split into a 7:3 training to testing set ratio. Data re-sampling was then applied to the training set. Lastly, a random forest classification algorithm was implemeted and validated on the testing data.](C:/Users/stili/Documents/TAD_predictive_modeling/Manuscript/Figures/figure3_ensemble_framework.png)

@@

![**Figure 4. SMOTE re-sampled models built on distance-type predictors yielded optimal MCCs across resolutions for GM12878.** Performances of random forest classification models built on each of the 3 predictor types: (A) OC, (B) OP, and (C) Distance. Within each predictor type, and across resolution, training data was re-sampled using different techniques including: None (red), ROS (green), RUS (blue), and SMOTE (purple). Performances were aggregated using means across each chromosome, with error bars representing 1 standard deviation away from the mean.](C:/Users/stili/Documents/TAD_predictive_modeling/Manuscript/Figures/figure4_gm12878_model_performances.png)

@@

![**Figure 5. The proportional importance of optimally built predictive models was outweighed by transcription factor binding sites across resolutions for GM12878.** Stacked barplots illustrate the proportion of total predictive importance in models using optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors), for each resolution on GM12878. Proportions are grouped by the class of cell line specific annotations in the feature space including chromatin state segmentations (CSS), histone modifications (HM), and transcription factor binding sites (TFBS).](C:/Users/stili/Documents/TAD_predictive_modeling/Manuscript/Figures/figure5_proportion_of_importances_gm12878.png)

@@

![**Figure 6. Optimally built TFBS-specific models outperformed other annotation-specific models across resolution for GM12878** Performances were compared for random forest models built using specific groups of annotations as the feature space including: histone modifications (HM; red), chromatin state segmentations (CSS; green), transcription factor binding sites (TFBS; blue), and a feature space containing all 3 types combined (combined; purple). Each bar represents a model ran through the ensemble framework, with re-sampling technique and predictor-type characteristics fixed to SMOTE and distance-types. The error bars represent $\pm$ 1 standard deviation from the mean MCC across each chromosome.](C:/Users/stili/Documents/TAD_predictive_modeling/Manuscript/Figures/figure6_comparing_sets_of_annotations_gm12878.png)

@@

![**Figure 7. Top 20 most predictive transcription factor binding sites sorted by 10 kb for GM12878.** A heatmap illustrating the importance of the key cell line specific transcription factors for optimally built random forest models across resolution. Importances were sorted according to the 10 kb resolution and rescaled to values between 0 and 100. Transcription factors with higher values of predictive importance are denoted in dark red.](C:/Users/stili/Documents/TAD_predictive_modeling/Manuscript/Figures/figure7_variable_importance_heatmaps_tfbs_gm12878.png)

@@

![**Figure 8. Top predictive transcription factor binding sites were significantly closer to bins that overlapped with 10 kb flanked TAD boundaries on GM12878** Boxplots illustrating the distribution of distances (on $log_{2}$ scale) around 10 kb regions on the linear genome that either overlapped with a flanked TAD boundary (Yes) or did not (No), for the most influential cell line specific TFBS including SMC3, CTCF, RAD21, ZNF143, and YY1sc281. P-values obtained from Wilcoxon Rank Sum tests are provided assessing the significance [$NS$ = p > 0.05; $*$ = 0.01 < p $\le$ 0.05; $**$ = 0.001 < p $\le$ 0.01; $***$ = 0.0001 < p $\le$ 0.001; $****$ = p $\le$ 0.00001].](C:/Users/stili/Documents/TAD_predictive_modeling/Manuscript/Figures/figure8_distance_boxplots_key_tfbs_gm12878_10kb.png)

>>>>

To investigate class imbalance, defined as the ratio of TAD vs. non-TAD boundaries, we estimated the number of TAD boundaries ($Y=1$ in the response vector) and the rest of genomic bins ($Y=0$) at four resolutions of Hi-C data. Expectedly, the average number of TAD boundaries and the rest of genomic bins increased with the increased resolution of Hi-C data (Table 1) irrespectively of cell line (Supplementary Tables 3). <!--This correlation [?? What correlation ??] with resolution was found to be significant, as evidenced by large intraclass correlation coefficients (ICC) (Table 2; Supplementary Tables 4). However, the class imbalance problem, caused by the sparsity of TAD boundaries throughout the genome, was found to be much more consistant across resolutions in both cell lines. This imbalance was evidenced by low imbalance ratios, defined as the ratio of the number of genomic bins that overlapped with flanked TAD boundaries, to the much larger number of genomic bins that did not. This relationship was not significantly correlated with resoltuion for either cell line, indicating that there existed a severe disparity between classes at all resolutions of Hi-C data.-->

**Table 1. TAD boundaries are the minority of the genome, creating high class imbalance ratio irrespectively of resolution.**

@@

<!--**Table 2. Inter-Resolution ICCs.** For GM12878, two-way intraclass correlation coefficients between resolutions for the following quantities: Number of genomic bins, number of TADs called, number of unique TAD boundaries, and imbalance ratio. The ICC was computed across 21 of the autosomal chromosomes. P-values indicate the level of significance of the correlation between data resolution and each quantity.[??? What is ICC? !!!editted]

| Quantity                 | Inter-Resolution | 95% CI          | P-Value  |
|--------------------------|------------------|-----------------|----------|
| Number of Genomic Bins   | 0.464            | [0.244, 0.690]  | < 0.0001 |
| Number of TADs Called    | 0.627            | [0.426, 0.802]  | < 0.0001 |
| Number of TAD Boundaries | 0.611            | [0.407, 0.791]  | < 0.0001 |
| Imbalanced Ratio         | 0.114            | [-0.058, 0.372] | 0.109    |
-->
<!--
## HIGH JACCARD INDICES IMPLY COLOCALIZATION OF MANY GROUP-SPECIFC ANNOTATIONS 
-->

## Three types of associations between TAD boundaries and epigenomic features

<!-- of TAD boundaries were assessed using functional genomic annotations, derived by ChIP-seq peak algorithms. Specific groups of annotations were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] and included chromatin state segementations, histone modifications, and transcription factor binding sites (TFBS).-->

@@

To establish the feature space for classification modeling, we constructed three different predictor types. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association between genomic elements (Figure 2A). **Overlap percent** (OP) represents a more precise measure of associations as it considers the proportion of a bin overlapped by a genomic annotation (Figure 2B). **Distance** between the centers of bins and genomic annotations quantifying the spatial relationship between genomic elements is the least common type of association (Figure 2C). Log2-transformation of distance was used as a normalization technique in order to account for the skewness present in the distributions of distances (Supplementary Figures 1-2). These types of genomic associations provide a granular way to investigate classification performance in class imbalance settings.

![**Three types of associations between TAD boundaries and genomic annotations.** (A) _Overlap Counts_ (OC) consider the total number of genomic features that overlapped with each genomic bin. (B) _Overlap Percent_ (OP) was calculated by dividing the sum of all feature widths within each specific bin by the total bin width. (C) _Distance_ ($log_{2}$ transformed) was calculated by measuring the distance (in bases) from the center of each genomic bin to the center of the nearest region defining a genomic annotation. ](Figures/figure2_predictor_types.png)

@@

![**Figure 3. Ensemble Framework.** A diagram of the cell line specific ensemble framework detailing the combination of inputs used in the model building pipeline. The minoirty and majority classes were represented by the genomic bins that overlapped with flanked TAD boundaries or did not, respectively. Functional genomic annotations obtained from the UCSC Genome Browser Database composed the feature space ($\textbf{X} = \{X_1, X_2, ..., X_p\}$). A feature engineering step was included to isolate the predictor-type specific spatial relationship between annotations and genomic bins in increasing granularity, denoted by the red gradient. The data was split into a 7:3 training to testing set ratio. Data re-sampling was then applied to the training set. Lastly, a random forest classification algorithm was implemeted and validated on the testing data.](Figures/figure3_ensemble_framework.png)

@@

![**Figure 4. SMOTE re-sampled models built on distance-type predictors yielded optimal MCCs across resolutions for GM12878.** Performances of random forest classification models built on each of the 3 predictor types: (A) OC, (B) OP, and (C) Distance. Within each predictor type, and across resolution, training data was re-sampled using different techniques including: None (red), ROS (green), RUS (blue), and SMOTE (purple). Performances were aggregated using means across each chromosome, with error bars representing 1 standard deviation away from the mean.](Figures/figure4_gm12878_model_performances.png)

@@

![**Figure 5. The proportional importance of optimally built predictive models was outweighed by transcription factor binding sites across resolutions for GM12878.** Stacked barplots illustrate the proportion of total predictive importance in models using optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors), for each resolution on GM12878. Proportions are grouped by the class of cell line specific annotations in the feature space including chromatin state segmentations (CSS), histone modifications (HM), and transcription factor binding sites (TFBS).](Figures/figure5_proportion_of_importances_gm12878.png)

@@

![**Figure 6. Optimally built TFBS-specific models outperformed other annotation-specific models across resolution for GM12878** Performances were compared for random forest models built using specific groups of annotations as the feature space including: histone modifications (HM; red), chromatin state segmentations (CSS; green), transcription factor binding sites (TFBS; blue), and a feature space containing all 3 types combined (combined; purple). Each bar represents a model ran through the ensemble framework, with re-sampling technique and predictor-type characteristics fixed to SMOTE and distance-types. The error bars represent $\pm$ 1 standard deviation from the mean MCC across each chromosome.](Figures/figure6_comparing_sets_of_annotations_gm12878.png)

@@

![**Figure 7. Top 20 most predictive transcription factor binding sites sorted by 10 kb for GM12878.** A heatmap illustrating the importance of the key cell line specific transcription factors for optimally built random forest models across resolution. Importances were sorted according to the 10 kb resolution and rescaled to values between 0 and 100. Transcription factors with higher values of predictive importance are denoted in dark red.](Figures/figure7_variable_importance_heatmaps_tfbs_gm12878.png)

@@

![**Figure 8. Top predictive transcription factor binding sites were significantly closer to bins that overlapped with 10 kb flanked TAD boundaries on GM12878** Boxplots illustrating the distribution of distances (on $log_{2}$ scale) around 10 kb regions on the linear genome that either overlapped with a flanked TAD boundary (Yes) or did not (No), for the most influential cell line specific TFBS including SMC3, CTCF, RAD21, ZNF143, and YY1sc281. P-values obtained from Wilcoxon Rank Sum tests are provided assessing the significance [$NS$ = p > 0.05; $*$ = 0.01 < p $\le$ 0.05; $**$ = 0.001 < p $\le$ 0.01; $***$ = 0.0001 < p $\le$ 0.001; $****$ = p $\le$ 0.00001].](Figures/figure8_distance_boxplots_key_tfbs_gm12878_10kb.png)
